---
title: "Sharing VPR Data"
author: "E. O'Grady"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sharing VPR Data}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::knitr}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup}
library(vprr)
library(dplyr)
library(rjson)
#library(readxl)
```

# Introduction
This document describes the process of saving and exporting VPR data in a format
which is easy to produce, and meets international data expectations of being, 
open source and accessible. It will describe the process of saving data outputs,
cover recommended formats, metadata and data naming suggestions, and more. Note 
that this is specific to data produced and shared by researchers at Bedford 
Institute of Oceanography (BIO), but the references may be helpful to a wider
audience. 


# Core Concepts
- FAIR
  A set of modern data sharing principles published by Wilkinson, et.al in 2016. 
  Findable, Accessible, Interoperable, Reproducible. 
  These principles mean that data should be shared widely and in formats which
  are straight-forward for others to access. These principles guide all the 
  decisions made in this vignette.
- Open Source formats
  Open source file formats do not require proprietary software in order to
  download, open or manipulate the files
- DarwinCore
  A data and metadata vocabulary which sets standard names for variables 
  relevant to biological diversity. (https://dwc.tdwg.org/)
- British Oceanographic Data Centre Vocabularies (BODC::P01)
  A data vocabulary which sets standard names for oceanographic variables. 
  (https://www.bodc.ac.uk/resources/vocabularies/parameter_codes/)
  

# Data products
VPR data can be represented in many forms. In this document we will focus on two
formats of final data products, which include environmental & abundance data, in
wide and long tables. In both forms, abundance has been calculated by
classification category and some relevant metadata has been added. 

The long format of data looks like
```{r}
data(category_conc_n)
head(category_conc_n)
```

The wide format looks like
```{r}
data(ctd_roi_merge)
head(ctd_roi_merge)
```

# Reccommended formats
During data processing and for internal work purposes, .Rdata files make a 
great storage option. They are easy to export with `save()`. You can capture 
the data in a variety of formats but most relevant might be a data frame or
an `oce` object. The advantage of the `oce` object is that it contains a 
metadata slot which can be useful for preserving self-contained metadata. 

When considering sharing the data, more accessible and generic formats should be
considered. During a review of file format options, researchers at BIO have
decided to utilize a combination of .csv files (for data storage) and .json
files (for metadata storage). This combination of files has the advantage of 
maintaining very simple file formats while capturing a wealth of associated 
metadata. The files are both very portable and easily accessed through a 
simple text editor, by users who are not data-experts.


# Metadata conventions
Metadata can be stored at a data set level, representing a 'cruise' or series of 
VPR deployments, as well as at an event (station, or cast) level, representing a
single VPR deployment. Each level has unique requirements but it should be
ensured that the metadata files have a connection (in this example: dataset_ID).

```{r}
# vpr_metadata_template <- system.file("extdata", "vpr_metadata_template.xlsx",
#                                      package = "vprr")
# 
# metadata_temp_ds <- readxl::read_xlsx(vpr_metadata_template, sheet = 1)
# 
# metadata_temp_st <- readxl::read_xlsx(vpr_metadata_template, sheet = 2)

```

The recommended format for preserving metadata is JSON A simple, easy to read,
file with a well documented structure. To generate a json file from R is very 
straightforward and can be done through `rjson::toJSON()`. This function will 
accept a nested list object and format it into a JSON string which can then be 
written to an external file. 

# Data conventions
Columns in VPR data outputs (from `vprr`) should be renamed for standardization.
This example uses a combination of DarwinCore (DwC) and British Oceanographic 
Data Centre (BODC) P01 vocabularies. Since VPR data contains a unique mix of 
biological and oceanographic environmental data, it was necessary to use both 
vocabularies to capture the full scope of the data. Data column names, as well
as units and definitions should be included in the metadata (shown in the 
example below in the 'dataAttributes' field). 


# Example using package data 

```{r}
data(category_conc_n)
metadata <- list("station_level" = list(
     "title" = list("en" = "VPR data from the Scotian Shelf",
                     "fr" = "Données VPR de l'étagère néo-écossaise"),
     "dataset_ID" = 1,
     "decimalLatitudeStart" = 44.5,
     "decimalLongitudeStart" = -64.5,
     "decimalLatitudeEnd" = 45.5,
     "decimalLongitudeEnd" = -65.5,
     "maximumDepthInMeters" = 1000,
     "eventDate" = "2019-08-11",
     "eventTime" = "00:00:00",
     "basisOfRecord" = "MachineObservation",
     "associatedMedia" = "https://ecotaxa.obs-vlfr.fr/ipt/archive.do?r=iml2018051",
     "identificationReferences" = "Iv3 model v3.3",
     "instrument" = list( "opticalSetting" = "S2",
                           "imageVolume" = 83663),
     "resources" = list("data" = list("name" = "vpr123_station25.csv",
                         "creationDate" = "2023-01-01"),
                         "metadata" = list("name" = "vpr123_station25-metadata.json",
                         "creationDate" = "2023-01-01")),
     "dataAttributes" = list("eventID" = list(
                                   "dataType" = "chr",
                                   "definition" = "An identifier for the set of information associated with a dwc:Event (something that occurs at a place and time). May be a global unique identifier or an identifier specific to the data set.",
                                    "vocabulary" = "dwc"),
                              "minimumDepthInMeters" = list(
                                    "dataType" = "float",
                                    "definition" = "The lesser depth of a range of depth below the local",
                                    "vocabulary" = "dwc"),
                              "maximumDepthInMeters" = list(
                                    "dataType" = "float",
                                    "definition" = "The greater depth of a range of depth below the local",
                                    "vocabulary" = "dwc"),
                              "DEPHPRST" = list(
                                    "dataType" = "float",
                                    "definition" = "Depth (spatial coordinate) of sampling event start relative to water surface in the water body by profiling pressure sensor and conversion to depth using unspecified algorithm",
                                    "vocabulary" = "BODC::P01"),
                               "individualCount" = list(
                                    "dataType" = "float",
                                    "definition" = "The number of individuals present at the time of the dwc:Occurrence.",
                                    "vocabulary" = "dwc"),
                               "verbatimIdentification" = list(
                                    "dataType" = "chr",
                                    "definition" = "A string representing the taxonomic identification as it appeared in the original record.",
                                    "vocabulary" = "dwc"),
                               "SDBIOL01" = list(
                                    "dataType" = "float",
                                    "definition" = "Abundance of biological entity specified elsewhere per unit volume of the water body",
                                    "vocabulary" = "BODC::P01"),
                               "TEMPST01" = list(
                                    "dataType" = "float",
                                    "definition" = "Temperature of the water body by CTD or STD",
                                    "vocabulary" = "BODC::P01"),
                               "PSALST01" = list(
                                    "dataType" = "float",
                                    "definition" = "Practical salinity of the water body by CTD and computation using UNESCO 1983 algorithm",
                                    "vocabulary" = "BODC::P01"),
                               "POTDENS0" = list(
                                    "dataType" = "float",
                                    "definition" = "Density (potential) of the water body by computation from salinity and potential temperature using UNESCO algorithm with 0 decibar reference pressure",
                                    "vocabulary" = "BODC::P01"),
                               "FLUOZZZZ" = list(
                                    "dataType" = "float",
                                    "definition" = "Fluorescence of the water body",
                                    "vocabulary" = "BODC::P01"),
                               "TURBXXXX" = list(
                                    "dataType" = "float",
                                    "definition" = "Turbidity of water in the water body",
                                    "vocabulary" = "BODC::P01"),
                               "sampleSizeValue" = list(
                                    "dataType" = "float",
                                    "definition" = "A numeric value for a measurement of the size (time duration, length, area, or volume) of a sample in a sampling dwc:Event.",
                                    "vocabulary" = "dwc"),
                               "sampleSizeUnit" = list(
                                    "dataType" = "chr",
                                    "definition" = "The unit of measurement of the size (time duration, length, area, or volume) of a sample in a sampling dwc:Event.",
                                    "vocabulary" = "dwc"),
                               "scientificName" = list(
                                    "dataType" = "chr",
                                    "definition" = "The full scientific name, with authorship and date information if known. When forming part of a dwc:Identification, this should be the name in lowest level taxonomic rank that can be determined. This term should not contain identification qualifications, which should instead be supplied in the dwc:identificationQualifier term.",
                                    "vocabulary" = "dwc"),
                               "identifiedBy" = list(
                                    "dataType" = "chr",
                                    "definition" = "A list (concatenated and separated) of names of people, groups, or organisations who assigned the Taxon to the subject.",
                                    "vocabulary" = "dwc"),
                               "identificationVerificationStatus" = list(
                                    "dataType" = "chr",
                                    "definition" = "A categorical indicator of the extent to which the taxonomic identification has been verified to be correct.",
                                    "vocabulary" = "dwc"),
                               "depthDifferenceMeters" = list(
                                    "dataType" = "float",
                                    "definition" = "Difference between maximumDepthInMeters and minimumDepthInMeters of an individual data bin, in meters",
                                    "vocabulary" = "BIO"),
                               "minimumTimeSeconds" = list(
                                    "dataType" = "float",
                                    "definition" = "minimum time value in a data bin, measured in seconds from the start of the day of sampling",
                                    "vocabulary" = "BIO"),
                               "maximumTimeSeconds" = list(
                                    "dataType" = "float",
                                    "definition" = "maximum time value in a data bin, measured in seconds from the start of the day of sampling",
                                    "vocabulary" = "BIO"),
                               "timeDifferenceSeconds" = list(
                                    "dataType" = "float",
                                    "definition" = "Difference between maximumTimeSeconds and minimumTimeSeconds of an individual data bin, in seconds",
                                    "vocabulary" = "BIO"),
                               "numberOfFrames" = list(
                                    "dataType" = "float",
                                    "definition" = "number of VPR frames captured within an individual data bin",
                                    "vocabulary" = "BIO"),
                               "timeMilliseconds" = list(
                                    "dataType" = "float",
                                    "definition" = "Time measured in milliseconds since the start of the sampling day",
                                    "vocabulary" = "BIO"),
                               "towyoID" = list(
                                    "dataType" = "chr",
                                    "definition" = "A string identifying the section of the cast to which the data point belongs",
                                    "vocabulary" = "BIO"),
                               "maximumCastDepthInMeters" = list(
                                    "dataType" = "float",
                                    "definition" = "Maximum depth in Meters of the cast dataset",
                                    "vocabulary" = "BIO")
   )))

   # new_name = old_name
columnNames = list( "DEPHPRST" = "depth" ,
   "verbatimIdentification" = "category",
   "eventID" = "station",
   "minimumDepthInMeters" = "min_depth",
   "maximumDepthInMeters" = "max_depth",
   "individualCount" = "n_roi_bin",
   "SDBIOL01" = "conc_m3",
   "TEMPST01" = "temperature",
   "PSALST01" = "salinity",
   "POTDENS0" = "density",
   "FLUOZZZZ" = "fluorescence",
   "TURBXXXX" = "turbidity",
   "sampleSizeValue" = "vol_sampled_bin_m3",
   "depthDifferenceMeters" = "depth_diff",
   "minimumTimeSeconds" = "min_time_s",
   "maximumTimeSeconds" = "max_time_s",
   "timeDifferenceSeconds" = "time_diff_s",
   "numberOfFrames" = "n_frames",
   "timeMilliseconds" = "time_ms",
   "towyoID" = "towyo",
   "maximumCastDepthInMeters" = "max_cast_depth"
   )

# add any new data columns required
# (eg. sampleSizeUnit, scientificName, identifiedBy, identificationVerificationStatus)
sampleSizeUnit <- "cubic metre"
identifiedBy <- "K. Sorochan"
identificationVerificationStatus <- "ValidatedByHuman"

data <- category_conc_n %>%
  dplyr::mutate(., identifiedBy = identifiedBy,
   sampleSizeUnit = sampleSizeUnit,
   identificationVerificationStatus = identificationVerificationStatus)

 # Define the mapping between category and scientific name
 # scientific names based ecotaxa taxonomic system
scientificName <- list("blurry" = "bad_image_blurry",
                       "artefact" = c("bad_image_malfunction", "bad_image_strobe"),
                       "Calanus" = "Calanus")

# Create a new column of data called scientificName based on matches to category
data <- data %>%
  dplyr::mutate(., scientificName = case_when(
    category %in% scientificName[["blurry"]] ~ "blurry",
    category %in% scientificName[["artefact"]] ~ "artefact",
    category == scientificName[["Calanus"]] ~ "Calanus",
    TRUE ~ NA
  ))

# vpr_export(data, metadata, columnNames, file = "vpr123_station25")

```

# Example of outputs
Here are the outputs from the above example.

```{r}
data <- read.csv(system.file(
  "extdata", "vpr123_station25.csv", package = "vprr"))
head(data)

metadata <- rjson::fromJSON(file = system.file(
  "extdata", "vpr123_station25-metadata.json", package = "vprr"))
```


# Conclusion
This document has summarized some of the reasoning behind creating and sharing
generic, easily accessible versions of VPR data products, as well as some 
methodology which could be used to produce standardized data and metadata 
products. Hopefully this will encourage wide spread sharing of VPR data products
and help future research aims.


# References

Wilkinson, M., Dumontier, M., Aalbersberg, I. et al. The FAIR Guiding Principles for scientific data management and stewardship. Sci Data 3, 160018 (2016). https://doi.org/10.1038/sdata.2016.18
